RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-12-01 manuscripts/manuscript.Rmd
Edits the Writing NeXML section.
Most of these are fairly minor. I did separate the explanation for creating
`nexml` objects into its own subsection, for better consistency with the
part about reading NeXML files.

## Writing phylogenetic trees

Generating NeXML files from R is likewise straightforward.  Here we
write a phylogeny from the `ape::phylo` format out to a NeXML file:

@@

The first argument to `nexml_write()` is a `nexml` object, or any
object that can be coerced to it, such as a `ape::phylo` phylogeny,
or a `multiPhylo` list of phylogenies. The `nexml_write()` function
also allows to add character data:

@@

These examples use data from the popular phylogenetics software `geiger` [@Harmon_2008].
Instead of packaging trees and characters at the time of writing a
NeXML file, the document can also be built up step by step using the
functions `add_characters()`,
`add_trees()`, and `add_meta()`.  Each of these functions can add data to an
existing `nexml` object, or create a new nexml object if none is provided.
Here we first create a `nexml` object containing the phylogeny data,
and then add the character data to it:

```{r}
geiger_nex <- add_trees(geospiza$phy)
geiger_nex <- add_characters(geospiza$dat, geiger_nex)
```

The data need not share the same taxa.  We can append additional
phylogenies and characters corresponding to different taxa:

```{r}
data(primates)
geiger_nex <- add_trees(primates$phy, geiger_nex)
geiger_nex <- add_characters(primates$dat, geiger_nex)
```

This allows us to include multiple trees and character matrices to a
single NeXML file, for instance, to represent samples from posterior
probability distributions of inferred or simulated phylogenies or
character states.

>>>>

## Writing phylogenetic trees and character data

Aside from providing an easy-to-use interface to data already in
NeXML, `RNeXML` also aims to facilitate broader adoption of the
standard for sharing phylogenetic data, and thus to make it
straightforward to generate them. The method `nexml_write()`, as the
name suggests, generates a NeXML file from its input parameters. In
its simplest invocation, the method writes a tree to a file:

@@

The first argument to `nexml_write()` is an object of type `nexml`, or any
object that can be coerced to it, such as in the above example an
`ape::phylo` phylogeny. Alternatively, passing a `multiPhylo` object
would write a list of phylogenies to the file.

In addition to trees, the `nexml_write()` method also to specify
character data as another parameter. The following example uses data
from the comparative phylogenetics R package `geiger` [@Harmon_2008].

@@

## Creating and populating `nexml` objects

Instead of packaging the various components for a NeXML file at the
time of writing the file, the user can also create and iteratively
populate an in-memory R object representation of the NeXML document,
i.e., an object of type `nexml`. The methods to do this are `add_characters()`,
`add_trees()`, and `add_meta()`, for adding characters, trees, and
metadata, respectively. Each of these functions will automatically
create a new nexml object if not supplied with an existing one as the
last (optional) argument.

For example, here we use `add_trees()` to first create a `nexml`
object with the phylogeny data, and then add the character data to it:

```{r}
nexObj <- add_trees(geospiza$phy)
nexObj <- add_characters(geospiza$dat, nexObj)
```

The data components being added to a `nexml` object need not share the
same OTUs.  For example, here we append additional phylogenies and characters
using different OTUs to the `nexml` object originally populated with
the `geospiza` phylogeny and character data:

```{r}
data(primates)
nexObj <- add_trees(primates$phy, nexObj)
nexObj <- add_characters(primates$dat, nexObj)
```

<!-- Might be worth adding some brief explanation here as to what -->
<!-- happens internally in such a case. For example, automatic -->
<!-- generation of multiple `otus` blocks? HL --> 
Other than storage size, there is no limit to the number of
phylogenies and character matrices that can be included in a
single NeXML file. This allows, for example, to capture samples from a
posterior probability distribution of inferred or simulated phylogenies or
character states in one NeXML file.
